---
title: "Data processing"
format: html
---

### Load the libraries

### s253699

```{r}
library("tidyverse")
library("here")
```

```{r}
ggplot(borovecki_tibble |>
  filter(gene %in% c("HTT","FAN1", "LIG1", "MLH1", "MSH3", "PMS1", "PMS2")), aes(gene, expressionValue, fill = diagnosis)) +
  geom_boxplot() +
  theme_minimal()
  

```

```{r}
ggplot(borovecki_tibble |>
filter(gene == "HTT"), aes(diagnosis, expressionValue, fill = diagnosis)) +
  geom_violin() +
  theme_minimal()

```

For HTT gene, symptomatic patientes tend to show a lower gene expression level compared to control patients. This is particularly interesting, because HTT is the key gene involved in Huntington's disease.
https://www.uniprot.org/uniprotkb/P42858/entry

```{r}
ggplot(borovecki_tibble |>
filter(gene == "PMS1"), aes(diagnosis, expressionValue, fill = diagnosis)) +
  geom_violin() +
  theme_minimal()
```

For PMS1, symptomatic patients show a higher expression level compared to controls.

```{r}
ggplot(borovecki_tibble |>
filter(gene == "ZDHHC17"), aes(diagnosis, expressionValue, fill = diagnosis)) +
  geom_violin() +
  theme_minimal()
```

For ZDHHC17, there is a higher gene expression for symptomatic patients compared to control patients. ZDHHC17 is a HTT-interacting protein that plays a key role in palmitoylation of neuronal proteins. https://www.uniprot.org/uniprotkb/Q8IUH5/entry
