---
title: "Data processing"
format: html
---

### Load libraries and data

```{r, echo=FALSE}
library("tidyverse")
library("fs")
library("patchwork")

borovecki_data <- read_rds(file = "/net/pupil1/home/people/s253704/R4BDS/projects/groupal_project/data/borovecki_tibble.rds")
```

### Data overview and inspections

We load the data to get an overview

```{r}
borovecki_data
```

### Classification according to expression level

As this is microRNA data, we will not get negative values, since we are measuring expression levels. If value is 0, it means there is no expression at all.

We calculate log2 to normalize the data. In this situation, we will have negative values, corresponding to the lowest values in the microRNA data matrix.

```{r}
borovecki_data <- borovecki_data |> 
  mutate(log2_expression = log2(expressionValue))
```

We need to know the min and max values in log2_expression to group the genes and study their expression levels.

```{r}
borovecki_data |>
  summarise(min_exp = min(log2_expression),
            max_exp = max(log2_expression))
```

In the next step, we created a categorical variable that groups gene expression values into six expression categories. Since the minimum and maximum observed expression levels in the dataset range from -3.3219 to 16.1313, we divided this continuous range into six intervals of equal width. Each interval represents a different level of gene expression, from the lowest category ("Very Low") to the highest one ("Very High").

```{r}
borovecki_grouped <- borovecki_data |> 
  mutate(expressionClass = case_when(log2_expression < -0.079717 ~ "Very Low",
    -0.079717 <= log2_expression & log2_expression < 3.162494 ~ "Low",
    3.162494 <= log2_expression & log2_expression < 6.404705 ~ "Moderate Low",
    6.404705 <= log2_expression & log2_expression < 9.646916 ~ "Moderate High",
    9.646916 <= log2_expression & log2_expression < 12.889127 ~ "High",
    12.889127 <= log2_expression ~ "Very High"))

borovecki_grouped |> 
  select(log2_expression, expressionClass)

```

We want to plot the different groups, to have an idea of the distribution in the data.

```{r}
borovecki_grouped |>
count(expressionClass) |>
  
# Visualization
ggplot(aes(x = expressionClass,
           y = n,
           fill = expressionClass)) +
  geom_col(colour = "green",
           fill = "lightgreen",
           alpha = 0.5) +
  theme_minimal(base_size = 16) +
  theme(axis.text.x = element_text(
        angle = 45,
        hjust = 1)) +
  labs(x = "Expression level group",
       y = "Number of genes",
       title = "Gene counts per expression category") +
  theme(legend.position = "none")
```

After grouping the expression values into six categories, we observed that most genes fall into the lower expression groups, while the uppermost group contains very few genes. This strongly suggests that the maximum value, 71,782.18, is likely an outlier.

In typical gene expression datasets, values usually show a right-skewed distribution, where a small number of genes have extremely high expression levels. The fact that almost all data points are concentrated in the first categories, and only a tiny fraction reaches the highest levels, supports the idea that the extreme value is not representative of the general expression distribution.

Specifically, we will see how many genes are in each expression level group.

```{r}
# Get total count in each expression group and percent.
borovecki_grouped |> 
  count(expressionClass) |> 
  mutate(percent = round(100 * n / sum(n), 2))

print(expression_counts)
```
