---
title: "Linear modeling"
format:
  html:
    embed-resources: true
editor: visual
---

## Load libraries

```{r}
library(tidyverse)  
library(modelr)     
library(splines)   
library(broom)      
library(ggrepel)  
```

## Load Data

```{r}
#| eval: true
borovecki_tibble <- readRDS("~/projects/group04_project/data/borovecki_tibble.rds")
```

## Clean Data

The `lm()` function uses a wide-data format where:

-   Each column is a gene

-   Each row is a sample denoted with its diagnosis: `0 = healthy`, `1 = disease`

-   Each cell contains the log2(expressionValue)

```{r}
borovecki_tibble <- borovecki_tibble |>
  pivot_wider(
    # Columns not to pivot
    id_cols = c(sample, diagnosis),
    
    # Columns to pivot
    names_from = gene,
    values_from = expressionValue
  ) |>
  
  # Create a new binary column that stores the clinical condition
  mutate(
    Diagnosis = case_when(
      diagnosis == "symptomatic" ~ 1,
      diagnosis == "control" ~ 0
    )
  ) |>
  dplyr::select(-c(diagnosis,sample)) |>
  relocate(Diagnosis, .before = 1)
```

## Individual Analysis

PMS1 was identified as a potential modifier gene in Hungington's disease, to study it, we applied a linear model (lm):

-   `(Intercept) = 56.82`: The average expression value of PMS1in healthy (0) patients

-   `Diagnosis = 33.17`: The average change in PMS1 expression when moving from the reference group (0) to the disease group (1).

```{r}
my_first_model <- lm(
  PMS1 ~ Diagnosis , 
  data = borovecki_tibble)

# Obtain coefficients of the model
coef(my_first_model)
```

I can also obtain these values by grouping by Diagnosis to compute the mean value of the gene PMS1

```{r}
borovecki_tibble |>
  group_by(Diagnosis) |>
  summarize(
    mu = mean(PMS1, na.rm = TRUE), # This works
  )
```

Both calculations show that PMS1 expression is up-regulated in the disease condition, it is interesting to see if this up-regulation is statistically significant at a level of significance of alfa = 0.05

To do so, we compute summary() within the linear model

```{r}
summary(my_first_model)
```

The gene is up-regulated in patients type 1 (positive slope)and the p-value is 0.00647 \< 0.05 which indicates that the effect is statistically significant at that alfa = 0.05.

## Complete Analysis for the 13041 genes

The complete analysis requires tidy data, to obtain it we will pivot the wide format

```{r}
borovecki_tibble_long <- borovecki_tibble  |> 
  pivot_longer(
    cols = -Diagnosis,
    names_to = "gene",
    values_to = "log2_expr_level"
  ) |>
  mutate(
    log2_expr_level = log2(log2_expr_level)
  )
```

Next, we will group the tidy data by `gene` and then add the `nest()` and `ungroup()` functions to your pipeline

```{r}
borovecki_tibble_long_nested <- borovecki_tibble_long |> 
  group_by(gene) |> 
  nest() |> 
  ungroup()
```

The result is a **`<tibble [13,039× 2]>`** where each row is a gene with a nested data frame that contains multiples rows with (diagnosis + log2_expr_level).

It's a tibble that has information of **13,039** genes for 2 variables (Diagnosis and log2Exp)

## Fitting models

Next, I want to fit a linear model to each gene, to do so, I have to let `R` know that I want to work per gene

```{r}
by_gene <- borovecki_tibble_long_nested |> 
  # The tibble is internally being group by the gene, works for certain functions
  # NEST() need groups
  group_by(gene)
```

I will now create a function that applies a linear model. Using `map()`, I'll apply this function to `borovecki_tibble_long_nested` *per* gene

-   The result is a **`<tibble [13,039× 3]>`** where each row is a gene with two nested data frames. The first one contains multiples rows with (diagnosis + log2_expr_level), whereas the second one contains the output of the linear function.

```{r}
# Define lm function
linear_model <- function(df) {
  lm(
  log2_expr_level ~ Diagnosis , 
  data = df)
}

# Apply lm function to each gene
borovecki_tibble_long_nested  <- borovecki_tibble_long_nested  |>  
  mutate(model_object = map(data, linear_model))
```

We will now re-order the results to obtain an understandable format:

1.  Create a `model_object_tidy` that contains tidy data about the estimation (p.value, estimates...)
2.  Unnest the result
3.  Unpack the `model_object_tidy`
4.  Select columns of interest
5.  Add a variable q.value that denotes the significancy of a gene

```{r}
tidy_data <- function(df) {
  df |>
  tidy(conf.int = TRUE,
       conf.level = 0.95)
  }

borovecki_estimates <- borovecki_tibble_long_nested |>
  mutate(model_object_tidy = map(model_object, tidy_data)) |>
  unnest(model_object_tidy) |>
  filter(term == "Diagnosis") |>
  dplyr::select(gene, p.value, estimate, conf.low, conf.high) |> 
  mutate(q.value=p.adjust(p.value),
          is_significant = case_when(
            q.value >= 0.05 ~ "no",
            q.value < 0.05 ~ "yes"
            )
         )

```
I will create a label for the top 10 potential genes with a formula that combines the estimate and the pvalue (Formula from System Biology class)

```{r}
borovecki_estimates <- borovecki_estimates |>
  mutate(
    score_fc = -log10(p.value) * sign(log2(estimate))
  )

top_10_genes <- borovecki_estimates |>
  arrange(desc(abs(score_fc))) |>
  slice_head(n = 10) |>
  pull(gene)

borovecki_estimates <- borovecki_estimates |>
  mutate(
    top10 = case_when(
      gene %in% top_10_genes ~ gene,
      TRUE ~ ""
    ))

```

Next, we save the object into our data folder

```{r}
saveRDS(borovecki_estimates, file = "~/projects/group04_project/data/borovecki_estimates.rds")
```

Let's now order the significant genes by their estimates and save the results

```{r}
ordered_borovecki <- borovecki_estimates |> 
  filter(is_significant == "yes") |>
  arrange(desc(abs(estimate)))

saveRDS(ordered_borovecki, file = "~/projects/group04_project/data/ordered_borovecki.rds")
```

Let's see, plot and save the most differentially expressed genes with a forest plot

-   The image have been saved with the function ggsave()

```{r}
#| message = FALSE
forest <- borovecki_estimates |> 
  filter(is_significant == "yes") |> 
  arrange(p.value) |> 
  slice_head(n = 30) |> 
  mutate(
    gene=fct_reorder(gene, estimate)
  ) |> 
  ggplot(aes(x=estimate, y=gene))+
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +   
  geom_point() +
  labs(
    title = "Genes Associated with hungington disease",
    x = "Estimates (95%CIs)",
    y = NULL
  ) +
  theme_minimal() 

forest

ggsave(
  filename = "~/projects/group04_project/results/forest.png", 
  plot = forest,                                     
  device = "png",                               
  dpi = 300   # Resolution
)
```

Let's see, plot and save the most deferentially expressed genes with a forest plot

```{r}
#| message = FALSE
volcano <- ggplot(borovecki_estimates, aes(x = estimate, y = -log10(p.value))) + 
  geom_point(aes(color=is_significant, label=top10), alpha=0.5, size=1) +
  geom_text_repel(aes(color=is_significant, label = top10), size = 2, max.overlaps = Inf) +
  labs(
    title = "Genes Associated with hungington disease",
    subtitle = "Genes highlighted in turquoise were significant after multiple testing correction",
    x = "Estimates",
    y = "-log10(p)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") 

volcano

ggsave(
  filename = "~/projects/group04_project/results/volcano.png", 
  plot = top10,                                     
  device = "png",                               
  dpi = 300   # Resolution
)
```
